\documentclass{ximera}

\begin{document}
	\author{Stitz-Zeager}
	\xmtitle{Exercises for Graphs of Sine and Cosine}{}

\mfpicnumber{1} \opengraphsfile{ExercisesforGraphsofSineandCosine} % mfpic settings added 

\begin{question}
In Exercises \ref{sinecosinegraphfirst} - \ref{sinecosinegraphlast}, graph one cycle of the given function.  State the period, amplitude, phase shift and vertical shift of the function.

\begin{problem} \label{sinecosinegraphfirst}
$f(t) = 3\sin(t)$ 
\end{problem}

\begin{problem}
$g(t) = \sin(3t)$
\end{problem}

\begin{problem}
$h(t)  = -2\cos(t)$
\end{problem}

\begin{problem}
$f(t)  = \cos \left( t - \frac{\pi}{2} \right)$
\end{problem}

\begin{problem}
$g(t)  = -\sin \left( t + \frac{\pi}{3} \right)$
\end{problem}

\begin{problem}
$h(t) = \sin(2t - \pi)$
\end{problem}

\begin{problem}
$f(t)  = -\frac{1}{3}\cos \left( \frac{1}{2}t + \frac{\pi}{3} \right)$
\end{problem}

\begin{problem}
$g(t) = \cos (3t - 2\pi) + 4$ 
\end{problem}
\begin{problem}
$h(t)  = \sin \left( -t - \frac{\pi}{4} \right) - 2$
\end{problem}

\begin{problem}
$f(t) = \frac{2}{3} \cos \left( \frac{\pi}{2} - 4t \right) + 1$
\end{problem}

\begin{problem}
$g(t) = -\frac{3}{2} \cos \left( 2t + \frac{\pi}{3} \right) - \frac{1}{2}$
\end{problem}


\begin{problem} \label{sinecosinegraphlast}
$h(t) = 4\sin (-2\pi t + \pi)$  
\end{problem}

\end{question}

\begin{question}
In Exercises \ref{fitsinecosinefirst} - \ref{fitsinecosinelast},  a sinusoid is graphed. Find a formula for the sinusoid in the form $S(t) = A \sin(\omega t + \phi) + B$ and $C(t) = A \cos(\omega t + \phi) + B$.  Select $\omega$ so  $\omega > 0$. Check your answer by graphing.

\begin{problem}   \label{fitsinecosinefirst}  
%$S(t) = 4 \sin \left(t + \frac{\pi}{4} \right)$, $C(t) = 4 \cos \left(t - \frac{\pi}{4} \right)$

\begin{mfpic}[16]{-5}{5}{-5}{5}
\tlabel[cc](5,-0.30){\scriptsize $t$}
\tlabel[cc](0.25,5){\scriptsize $y$}
\axes
\xmarks{-4 step 1 until 4}
\ymarks{-4 step 1 until 4}
\point[4pt]{(-3.927, 0), (-2.356, -4), (-0.7854, 0), (0.7854, 4), (2.356, 0)}
\tlabel[cc](-3, 0.5){ \scriptsize $\left( -\frac{5\pi}{4}, 0 \right)$}
\tlabel[cc](-2.356, -4.5){ \scriptsize $\left( -\frac{3\pi}{4}, -4 \right)$}
\gclear \tlabelrect(0.25, -0.6){ \scriptsize $\left( -\frac{\pi}{4}, 0 \right)$ \vphantom{$\frac{5}{2}$}}
\tlabel[cc](0.9, 4.5){\scriptsize $\left( \frac{\pi}{4}, 4 \right)$}
\tlabel[cc](3.25, 0.5){ \scriptsize $\left( \frac{3\pi}{4}, 4 \right)$}
\tlpointsep{4pt}
%\axislabels {x}{{$\frac{\pi}{2}$} 1.5708, {$\pi$} 3.1416, {$\frac{3\pi}{2}$} 4.7124, {$2\pi$} 6.2832}
%\axislabels {y}{{$-3$} -3, {$3$} 3}
\penwd{1.25pt}
\arrow \reverse \arrow \function{-5, 5,  0.1}{4*sin(x+(pi/4))}
\end{mfpic}   
\end{problem}


\begin{problem} 
% $S(t) = -3 \sin(t) + 3$, $C(t) = -3 \cos\left(t - \frac{\pi}{2}\right) + 3$

\begin{mfpic}[16]{-6}{6}{-1}{9}
\tlabel[cc](6,-0.30){\scriptsize $t$}
\tlabel[cc](0.25,9){\scriptsize $y$}
\axes
\xmarks{-5 step 1 until 5}
\ymarks{1 step 1 until 8}
\point[4pt]{(-4.712, 0), (-1.571, 6), (0,3), (1.571, 0), (4.712, 6)}
\tlabel[cc](-4.712, -0.75){\scriptsize $\left( -\frac{3\pi}{2}, 0 \right)$}
\tlabel[cc](-1.571, 6.5){\scriptsize $\left( -\frac{\pi}{2}, 6 \right)$}
\tlabel[cc](0.75, 3){ \scriptsize $\left(0, 3 \right)$}
\tlabel[cc](4.712, 6.5){\scriptsize $\left( \frac{3 \pi}{2}, 6 \right)$}
\tlabel[cc](1.571, -0.75){\scriptsize $\left( \frac{\pi}{2}, 0 \right)$}
\tlpointsep{4pt}
%\axislabels {x}{{$\frac{\pi}{2}$} 1.5708, {$\pi$} 3.1416, {$\frac{3\pi}{2}$} 4.7124, {$2\pi$} 6.2832}
%\axislabels {y}{{$-3$} -3, {$3$} 3}
\penwd{1.25pt}
\arrow \reverse \arrow \function{-6, 6,  0.1}{3- 3*sin(x)}
\end{mfpic}  
\end{problem}

\begin{problem}
%$S(t) = 3 \sin \left( 2t - \frac{\pi}{3} \right)$, $C(t) = 3 \cos \left( 2t - \frac{5\pi}{6} \right)$

\begin{mfpic}[15]{-6}{6}{-5}{5}
\tlabel[cc](6,-0.30){\scriptsize $t$}
\tlabel[cc](0.25,5){\scriptsize $y$}
\axes
\xmarks{-5 step 1 until 5}
\ymarks{-4 step 1 until 4}
\point[4pt]{(-4.974, 3), (-4.189, 0), (-3.403, -3), (-2.618, 0), (-1.832, 3), (-1.047, 0), (-0.2618, -3), (0.5236, 0), (1.309, 3), (2.094, 0), (2.8880, -3), (3.665, 0), (4.45, 3), (5.236, 0)}
\tlabel[cc](-4.974, 3.75){ \scriptsize $\left( -\frac{19 \pi}{12}, 3 \right)$}
\tlabel[cc](-3.403, -3.75){ \scriptsize $\left( -\frac{13 \pi}{12}, -3 \right)$}
\tlabel[cc] (-1.832, 3.75){ \scriptsize $\left( -\frac{7 \pi}{12}, 3 \right)$}
\gclear \tlabelrect(0, -3.75){ \scriptsize $\left( -\frac{\pi}{12}, -3 \right)$}
\tlabel[cc] (1.309, 3.75){ \scriptsize $\left( \frac{5 \pi}{12}, 3 \right)$}
\tlabel[cc](2.880, -3.75){ \scriptsize $\left( \frac{11 \pi}{12}, -3 \right)$}
\tlabel[cc] (4.45, 3.75){ \scriptsize $\left( \frac{17 \pi}{12}, 3 \right)$}

\tlpointsep{4pt}
%\axislabels {x}{{$\frac{\pi}{2}$} 1.5708, {$\pi$} 3.1416, {$\frac{3\pi}{2}$} 4.7124, {$2\pi$} 6.2832}
%\axislabels {y}{{$-3$} -3, {$3$} 3}
\penwd{1.25pt}
\arrow \reverse \arrow \function{-6, 5.5,  0.1}{3*sin((2*x)-(pi/3))}
\end{mfpic}  
\end{problem}


\begin{problem} \label{fitsinecosinelast} 
% $S(t) =  \frac{7}{2} \sin(\pi t) + \frac{1}{2}$, $C(t) = \frac{7}{2} \cos\left(\pi t \frac{\pi}{2} \right) + \frac{1}{2}$

\begin{mfpic}[22][15]{-5}{5}{-5}{5}
\tlabel[cc](5,-0.30){\scriptsize $t$}
\tlabel[cc](0.25,5){\scriptsize $y$}
\axes
\xmarks{-4 step 1 until 4}
\ymarks{-4 step 1 until 4}
\point[4pt]{ (-4, 0.5), (-3.5, 4), (-3, 0.5), (-2.5, -3), (-2, 0.5), (-1.5, 4), (-1, 0.5), (-0.5, -3), (0, 0.5), (0.5, 4), (1, 0.5), (1.5,-3), (2, 0.5), (2.5,4), (3, 0.5), (3.5,-3), (4, 0.5) }
%\tlabel[cc](-4.5, -3.5){\scriptsize $(-4.5, -3)$}
\tlabel[cc](-3.5, 4.5){\scriptsize $(-3.5, 4)$}
\tlabel[cc](-2.5, -3.5){\scriptsize $(-2.5, -3)$}
\tlabel[cc](-1.5, 4.5){\scriptsize $(-1.5, 4)$}
%\tlabel[cc](-0.5, -3.5){\scriptsize $(-0.5, -3)$}
\gclear \tlabelrect(0.75, 4.5){\scriptsize $(0.5, 4)$}
\tlabel[cc](1.5, -3.5){\scriptsize $(1.5, -3)$}
\tlabel[cc](2.5, 4.5){\scriptsize $(2.5, 4)$}
%\tlabel[cc](2.5, -3.5){\scriptsize $(2.5, -3)$}
%\tlabel[cc](4.5, 4.5){\scriptsize $(4.5, 4)$}

\tlpointsep{4pt}
%\axislabels {x}{{$\frac{\pi}{2}$} 1.5708, {$\pi$} 3.1416, {$\frac{3\pi}{2}$} 4.7124, {$2\pi$} 6.2832}
%\axislabels {y}{{$-3$} -3, {$3$} 3}
\penwd{1.25pt}
\arrow \reverse \arrow \function{-4.25, 4.25,  0.1}{(3.5*sin(pi*x))+0.5}
\end{mfpic}   

\end{problem}

\end{question}

\begin{problem} Use the graph of  $S(t) = 4 \sin(t)$ to graph each of the following functions. State the period of each.

\begin{enumerate}

\item $f(t) = | 4 \sin(t)|$

\item $g(t) = \sqrt{4 \sin(t)}$

\end{enumerate}

\end{problem}

\begin{question}
In Exercises \ref{exploregraphsfirst} - \ref{exploregraphslast}, use a graphing utility to help you and your classmates discuss the given questions.


\begin{problem} \label{exploregraphsfirst}
Graph $f(t) = \cos(3t) + \sin(t)$.  Is this function periodic?  If so, what is the period? 
\end{problem}

\begin{problem} 
Graph $f(t) = t \sin(t)$ along with  $y = \pm t$. What do you notice? 
\end{problem}

\begin{problem}
Graph $f(t) = \frac{\sin(t)}{t}$ along with  $y = \pm \frac{1}{t}$.   What do you notice?

\begin{enumerate}

\item What appears to be $\ds{\lim_{t \rightarrow \infty} f(t)}$?    Interpret your answer graphically.



\item  Use the Squeeze Theorem, Theorem \ref{squeezeth} from Section \ref{Sequences} to prove your claim.  



\textbf{HINT:}  Since $-1 \leq \sin(t) \leq 1$, for $t>0$, $-\frac{1}{t} \leq \frac{\sin(t)}{t} \leq \frac{1}{t}$ \ldots

\end{enumerate}

\end{problem}


\begin{problem}
Graph $f(t) = \cos\left(\frac{1}{t}\right)$.  

\begin{enumerate}

\item Investigate  $\ds{\lim_{t \rightarrow 0} f(t)}$.  Does $\ds{\lim_{t \rightarrow 0} f(t)}$ exist?  Why or why not?

\item  Determine $\ds{\lim_{t \rightarrow \infty} f(t)}$ and interpret your answer graphically.

\end{enumerate}
\end{problem}

\begin{problem}  \label{exploregraphslast}
Graph $f(t) = e^{-0.1t} \left( \cos(2t) + \sin(2t)\right)$ along with $y = \pm 2 \, e^{-0.1t}$. What do you notice?  

\begin{enumerate}

\item What appears to be $\ds{\lim_{t \rightarrow \infty} f(t)}$?   Interpret your answer graphically.

\item  Use the Squeeze Theorem, Theorem \ref{squeezeth} from Section \ref{Sequences} to prove your claim.  

\textbf{HINT:}  Since $-1 \leq \cos(2t) \leq 1$ and $-1 \leq \sin(2t) \leq 1$, $-2 \leq \cos(2t) + \sin(2t) \leq 2$.  Hence, $-2 e^{-0.1t} \leq e^{-0.1t} \left( \cos(2t) + \sin(2t)\right) \leq 2e^{-0.1t}$ \ldots

\end{enumerate}

\end{problem}

\end{question}

\begin{problem}
Show every constant function $f$ is periodic by explaining why $f(x + 117) = f(x)$ for all real numbers $x$. Then show that $f$ has no period by showing that you cannot find a \emph{smallest} number $p$ such that $f(x + p) = f(x)$ for all real numbers $x$.  


Said differently, show that $f(x + p) = f(x)$ for all real numbers $x$ for ALL values of $p > 0$, so no smallest value exists to satisfy the definition of `period'.

\end{problem}

\begin{problem}
The sounds we hear are made up of mechanical waves.  The note `A' above the note `middle C' is a sound wave with ordinary frequency $f = 440$ Hertz $= 440 \frac{\text{cycles}}{\text{second}}$.  Find a sinusoid which models this note, assuming that the amplitude is $1$ and the phase shift is $0$.
\end{problem}

\begin{problem}
The voltage $V$ in an alternating current source has amplitude $220 \sqrt{2}$ and ordinary frequency $f = 60$ Hertz.  Find a sinusoid which models this voltage.  Assume that the phase is $0$.
\end{problem}


\begin{problem} \label{heightlondoneye} 
The \href{http://en.wikipedia.org/wiki/London_Eye}{\underline{London Eye}} is a popular tourist attraction in London, England and is one of the largest Ferris Wheels in the world.  It has a diameter of 135 meters and makes one revolution (counter-clockwise) every 30 minutes.  It is constructed so that the lowest part of the Eye reaches ground level, enabling passengers to simply walk on to, and off of, the ride.  Find a sinsuoid which models the height $h$ of the passenger above the ground in meters $t$ minutes after they board the Eye at ground level.
\end{problem}

\begin{problem} \label{leftrightlondoneye} 
On page \pageref{equationsforcircularmotion} in Section \ref{cosinesinebeyond}, we found the $x$-coordinate of counter-clockwise motion on a circle of radius $r$ with angular frequency $\omega$ to be $x = r\cos(\omega t)$, where $t=0$ corresponds to the point $(r,0)$.  Suppose we are in the situation of Exercise \ref{heightlondoneye} above.  Find a sinsusoid which models the horizontal \textit{displacement} $x$ of the passenger from the center of the Eye in meters $t$ minutes after they board the Eye.  Here we take $x(t) > 0$ to mean the passenger is to the \textit{right} of the center, while $x(t) < 0$ means the passenger is to the \textit{left} of the center.
\end{problem}

\begin{problem}
In Exercise \ref{yoyotrick} in Section \ref{RadianMeasure}, we introduced the yo-yo trick `Around the World' in which a yo-yo is thrown so it sweeps out a vertical circle.  As in that exercise, suppose the yo-yo string is 28 inches and it completes one revolution in 3 seconds.  If the closest the yo-yo ever gets to the ground is 2 inches, find a sinsuoid which models the height $h$ of the yo-yo above the ground in inches $t$ seconds after it leaves its lowest point.
\end{problem}


\begin{problem}
Consider the pendulum below.  Ignoring air resistance, the angular displacement of the pendulum from the vertical position, $\theta$, can be modeled as a sinusoid.\footnote{Provided $\theta$ is kept `small.'  Carl remembers the `Rule of Thumb' as being $20^{\circ}$ or less.  Check with your friendly neighborhood physicist to make sure.}


\begin{center}

\begin{mfpic}[15]{-3}{3}{-5}{1}
\polyline{(0,0), (0,-5)}
\dashed \polyline{(0,0), (2.5, -4.33)}
\arrow \parafcn{275, 295, 5}{4*dir(t)}
\tlabel[cc](1.29, -4.83){$\theta$}
\hatchcolor[gray]{.7}
\lhatch \rect{(-3,0), (3,1)}
\fillcolor[gray]{.7} 
\gfill \circle{(0,-5),0.25}
\gfill \circle{(2.5, -4.33),0.20}
\penwd{1.025}
\circle{(0,-5),0.25}
\circle{(2.5, -4.33),0.25}
\rect{(-3,0), (3,1)}
\end{mfpic} 
\end{center}

The amplitude of the sinusoid is the same as the initial angular displacement, $\theta_{\text{\tiny $0$}}$, of the pendulum and the  period of the motion is given by

\[T = 2\pi \sqrt{\frac{\ell}{g}}\]

where $\ell$ is the length of the pendulum and $g$ is the acceleration due to gravity.

\begin{enumerate}

\item  Find a sinusoid which gives the angular displacement $\theta$ as a function of time, $t$. Arrange things so $\theta(0) = \theta_{\text{\tiny $0$}}$.

\item  In Exercise \ref{pendulumproblem} section \ref{RootRadicalFunctions}, you found the length of the pendulum needed in Jeff's antique Seth-Thomas clock to ensure the period of the pendulum is $\frac{1}{2}$ of a second. Assuming the initial displacement of the pendulum is $15^{\circ}$, find a sinusoid which models the displacement of the pendulum $\theta$ as a function of time, $t$, in seconds. 

\end{enumerate}
\end{problem}


\begin{problem}
The table below lists the average temperature of Lake Erie as measured in Cleveland, Ohio on the first of the month for each month during the years 1971 -- 2000.\footnote{See this website: \href{http://www.erh.noaa.gov/cle/climate/cle/normals/laketempcle.html}{\underline{http://www.erh.noaa.gov/cle/climate/cle/normals/laketempcle.html}.}}  For example,   $t=3$ represents the average of the temperatures recorded for Lake Erie on every March 1 for the years 1971 -- 2000.

\medskip

\small

\noindent \begin{tabular}{|l|r|r|r|r|r|r|r|r|r|r|r|r|} \hline
Month  & & & & & & & & & & & & \\
Number, $t$ & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12\\ 
\hline 
Temperature  & & & & & & & & & & & & \\
($^{\circ}$ F), $T$ & 36 & 33 & 34 & 38 & 47 & 57 & 67 & 74 & 73 & 67 & 56 & 46 \\ \hline
\end{tabular}

\normalsize


\begin{enumerate}

\item \label{LakeErieTempData} Using the techniques discussed in Example \ref{sinusoidsunlight}, fit a sinusoid to these data. 

\item  Graph your model along with the data set to judge the reasonableness of the fit.

\item Use the model from \ref{LakeErieTempData} to predict the average temperature recorded for Lake Erie on April $15^{\text{th}}$ and September $15^{\text{th}}$ during the years 1971--2000.\footnote{The computed average is $41^{\circ}$F for April $15^{\text{th}}$ and $71^{\circ}$F for September $15^{\text{th}}$.}

\item Compare your results to those obtained using a graphing utility.

\end{enumerate}
\end{problem}

\begin{problem}
The fraction of the moon illuminated at midnight Eastern Standard Time on the $t^{\text{th}}$ day of June, 2009 is given in the table below.\footnote{See this website: \href{http://www.usno.navy.mil/USNO/astronomical-applications/data-services/frac-moon-ill}{\underline{http://www.usno.navy.mil/USNO/astronomical-applications/data-services/frac-moon-ill}.}} 


\small

\noindent \begin{tabular}{|l|r|r|r|r|r|r|r|r|r|r|} \hline
Day of  & & & & & & & & & & \\
June, $t$ & 3 & 6 & 9 & 12 & 15 & 18 & 21 & 24 & 27 & 30\\ 
\hline 
Fraction  & & & & & & & & & & \\
Illuminated, $F$ & 0.81 & 0.98 & 0.98 & 0.83 & 0.57 & 0.27 & 0.04 & 0.03 & 0.26 & 0.58  \\ \hline
\end{tabular}

\normalsize



\begin{enumerate}

\item \label{MoonIllumination} Using the techniques discussed in Example \ref{sinusoidsunlight}, fit a sinusoid to these data.\footnote{You may want to plot the data before you find the phase shift.} 

\item  Graph your model along with the data set to judge the reasonableness of the fit.

\item Use the model from \ref{MoonIllumination} to predict the fraction of the moon illuminated on June 1, 2009. \footnote{The listed fraction is $0.62$.}

\item Compare your results to those obtained using a graphing utility.

\end{enumerate}

\end{problem}


\begin{problem} \label{graphdesmosregression}  
Use a graphing utility to graph $y = 8.36 \sin(-294.81t  - 26.53) + 12.06$.  (This is the regression model produced by desmos in Example \ref{sinusoidsunlight}.) Zoom in, as needed, until you start to see the wave-like nature of the graph.  Use Theorem \ref{sinusoidform} to determine a window which produces exactly one complete cycle of this sinusoid and check your answer graphically.
\end{problem}


\begin{problem}  \label{proofsinusoidformexercise} 
Use Theorem \ref{transformationsthm} to prove Theorem \ref{sinusoidform}.
\end{problem}


\begin{problem}
With the help of your classmates, research \href{http://en.wikipedia.org/wiki/Amplitude_modulation}{\underline{Amplitude Modulation}} and \href{http://en.wikipedia.org/wiki/Frequency_modulation}{\underline{Frequency Modulation}}.
\end{problem}

\begin{problem} 
What other things in the world might be roughly sinusoidal?  Look to see what models you can find for them and share your results with your class.
\end{problem}

\end{document}