\documentclass{ximera}

\begin{document}
	\author{Stitz-Zeager}
	\xmtitle{Exercises for The Circular Functions Sine and Cosine}{}

\mfpicnumber{1} \opengraphsfile{ExercisesforTheCircularFunctionsSineandCosine} % mfpic settings added 


\begin{question}
    
In Exercises \ref{valuefirst} - \ref{valuelast}, find the exact value of the cosine and sine of the given angle.

\begin{problem}\label{valuefirst}
    $\theta = 0$
\end{problem}

\begin{problem}
    $\theta = \frac{\pi}{4}$
\end{problem}

\begin{problem}
    $\theta = \frac{\pi}{3}$
\end{problem}

\begin{problem}
    $\theta = \frac{\pi}{2}$
\end{problem}



\begin{problem}
    $\theta = \frac{2\pi}{3}$
\end{problem}

\begin{problem}
    $\theta = \frac{3\pi}{4}$
\end{problem}

\begin{problem}
     $\theta = \pi$
\end{problem}

\begin{problem}
    $\theta = \frac{7\pi}{6}$
\end{problem}

\begin{problem}
    $\theta = \frac{5\pi}{4}$
\end{problem}

\begin{problem}
    $\theta = \frac{4\pi}{3}$
\end{problem}

\begin{problem}
    $\theta = \frac{3\pi}{2}$
\end{problem}

\begin{problem}
    $\theta = \frac{5\pi}{3}$
\end{problem}

\begin{problem}
    $\theta = \frac{7\pi}{4}$
\end{problem}

\begin{problem}
    $\theta = \frac{23\pi}{6}$
\end{problem}

\begin{problem}
    $\theta = -\frac{13\pi}{2}$
\end{problem}

\begin{problem}
     $\theta = -\frac{43\pi}{6}$
\end{problem}

\begin{problem}
    $\theta = -\frac{3\pi}{4}$
\end{problem}

\begin{problem}
     $\theta = -\frac{\pi}{6}$
\end{problem}

\begin{problem}
    $\theta = \frac{10\pi}{3}$
\end{problem}

\begin{problem} \label{valuelast}
    $\theta = 117\pi$
\end{problem}

\end{question}

\begin{question}
In Exercises \ref{solveforanglefirst} - \ref{solveforanglelast}, find all of the angles which satisfy the given equation.


\begin{problem}\label{solveforanglefirst}
    $\sin(\theta) = \frac{1}{2}$
\end{problem}

\begin{problem}
    $\cos(\theta) = -\frac{\sqrt{3}}{2}$
\end{problem}

\begin{problem}
    $\sin(\theta) = 0$
\end{problem}

\begin{problem}
    $\cos(\theta) = \frac{\sqrt{2}}{2}$
\end{problem}

\begin{problem}
    $\sin(\theta) = \frac{\sqrt{3}}{2}$
\end{problem}

\begin{problem}
    $\cos(\theta) = -1$ 
\end{problem}

\begin{problem}
    $\sin(\theta) = -1$
\end{problem}

\begin{problem}
    $\cos(\theta) = \frac{\sqrt{3}}{2}$
\end{problem}

\begin{problem}\label{solveforanglelast}
    $\cos(\theta) = -1.001$
\end{problem}

\end{question}

\begin{question}
    
In Exercises \ref{solvefortfirst} - \ref{solvefortlast}, solve the equation for $t$.  (See the remarks on page \ref{cosinesineequationsrealnumbers}.)


\begin{problem}\label{solvefortfirst}
    $\cos(t) = 0$
\end{problem}

\begin{problem}
    $\sin(t) = -\frac{\sqrt{2}}{2}$
\end{problem}

\begin{problem}
    $\cos(t) = 3$
\end{problem}

\begin{problem}
    $\sin(t) = -\frac{1}{2}$
\end{problem}

\begin{problem}
    $\cos(t) = \frac{1}{2}$
\end{problem}

\begin{problem}
    $\sin(t) = -2$
\end{problem}

\begin{problem}
    $\cos(t) = 1$
\end{problem}

\begin{problem}
     $\sin(t) = 1$
\end{problem}

\begin{problem}\label{solvefortlast}
    $\cos(t) = -\frac{\sqrt{2}}{2}$
\end{problem}


\end{question}

\begin{question}
    
In Exercises \ref{pointsfirst} - \ref{pointslast}, let $\theta$ be the angle in standard position whose terminal side contains the given point then compute $\cos(\theta)$ and $\sin(\theta)$.

\begin{problem}\label{pointsfirst} 
    $P(-7, 24)$
\end{problem}

\begin{problem}
    $Q(3, 4)$
\end{problem}

\begin{problem}
    $R(5, -9)$
\end{problem}

\begin{problem}\label{pointslast}
    $T(-2, -11)$
\end{problem}

\end{question}


\newpage

\begin{question}
    
In Exercises \ref{findthevaluefirst} - \ref{findthevaluelast}, use the results developed throughout the section to find the requested value.



\begin{problem}\label{findthevaluefirst}
    If $\sin(\theta) = -\frac{7}{25}$ with $\theta$ in Quadrant IV, what is $\cos(\theta)$?
\end{problem}

\begin{problem}
    If $\cos(\theta) = \frac{4}{9}$ with $\theta$ in Quadrant I, what is $\sin(\theta)$?
\end{problem}

\begin{problem}
    If $\sin(\theta) = \frac{5}{13}$ with $\theta$ in Quadrant II, what is $\cos(\theta)$?
\end{problem}

\begin{problem}
    If $\cos(\theta) = -\frac{2}{11}$ with $\theta$ in Quadrant III, what is $\sin(\theta)$?
\end{problem}

\begin{problem}
    If $\sin(\theta) = -\frac{2}{3}$ with $\theta$ in Quadrant III, what is $\cos(\theta)$?
\end{problem}

\begin{problem}
    If $\cos(\theta) = \frac{28}{53}$ with $\theta$ in Quadrant IV, what is $\sin(\theta)$?
\end{problem}

\begin{problem}
     If $\sin(\theta) = \frac{2\sqrt{5}}{5}$ and $\frac{\pi}{2} < \theta < \pi$, what is $\cos(\theta)$?
\end{problem}

\begin{problem}
     If $\cos(\theta) = \frac{\sqrt{10}}{10}$ and $2\pi < \theta < \frac{5\pi}{2}$, what is $\sin(\theta)$?
\end{problem}

\begin{problem}
    If $\sin(\theta) = -0.42$ and $\pi < \theta < \frac{3\pi}{2}$, what is  $\cos(\theta)$?
\end{problem}

\begin{problem}\label{findthevaluelast}
    If $\cos(\theta) = -0.98$ and $\frac{\pi}{2} < \theta < \pi$, what is $\sin(\theta)$?
\end{problem}


\end{question}

\begin{question}
    
In Exercises \ref{calculatorfirst} - \ref{calculatorlast}, use your calculator to approximate the given value to three decimal places.  Make sure your calculator is in the proper angle measurement mode!

\begin{problem}\label{calculatorfirst}
    $\sin(78.95^{\circ})$
\end{problem}

\begin{problem}
    $\cos(-2.01)$
\end{problem}


\begin{problem}
    $\sin(392.994)$
\end{problem}

\begin{problem}
    $\cos(207^{\circ})$
\end{problem}

\begin{problem}
    $\sin\left( \pi^{\circ} \right)$
\end{problem}

\begin{problem}\label{calculatorlast}
    $\cos(e)$
\end{problem}

\end{question}

\begin{question}
    
In Exercises \ref{decomposebasicsinecosinefirst} - \ref{decomposebasicsinecosinelast}, write the given function as a nontrivial decomposition of functions as directed.

\begin{problem}\label{decomposebasicsinecosinefirst}
    For $f(t) = 3t + \sin(2t)$, find functions $g$ and $h$ so that $f=g+h$.
\end{problem}

\begin{problem}
    For $f(\theta) = 3\cos(\theta) - \sin(4\theta)$, find functions $g$ and $h$ so that $f=g-h$.
\end{problem}

\begin{problem}
    For $f(t) = e^{-0.1t} \sin(3t)$, find functions $g$ and $h$ so that $f=gh$.
\end{problem}


\begin{problem}
    For $r(t) = \frac{\sin(t)}{t}$, find functions $f$ and $g$ so $r = \frac{f}{g}$.
\end{problem}

\begin{problem}\label{decomposebasicsinecosinelast}
    For $r(\theta) =\sqrt{3 \cos(\theta)}$, find functions $f$ and $g$ so $r = g \circ f$.
\end{problem}

\end{question}


\begin{problem}\label{sinearcexercise}

\begin{enumerate}
\setcounter{enumi}{\value{HW}}

For each function $S(t)$ listed below, compute the average rate of change over the indicated interval.\footnote{See Definition \ref{arc} in Section \ref{AverageRateofChange} for a review of this concept, as needed.}  What trends do you notice? Be sure your calculator is in radian mode!

\[ \begin{array}{|r||c|c|c|}  \hline

 S(t) &  [-0.1, 0.1] & [-0.01, 0.01] &[-0.001, 0.001] \\ \hline
 \sin(t)     &&&  \\  \hline
 \sin(2t)   &&&  \\ \hline
 \sin(3t)   &&&   \\  \hline
\sin(4t)   &&&   \\  \hline

\end{array} \]

\setcounter{HW}{\value{enumi}}
\end{enumerate}

\end{problem}


\begin{question}
    In Exercises \ref{motionfirst} - \ref{motionlast}, find the equations of motion for the given scenario.  Assume that the center of the motion is the origin, the motion is counter-clockwise and that $t = 0$ corresponds to a position along the positive $x$-axis.  (See Equation \ref{equationsforcircularmotion} and Example \ref{EarthRotationEx}.)

\begin{problem}\label{motionfirst} 
    A point on the edge of the spinning yo-yo in Exercise \ref{spinningyoyo} from Section \ref{RadianMeasure}. 

Recall: The diameter of the yo-yo is 2.25 inches and it spins at 4500 revolutions per minute.
\end{problem}

\begin{problem}
     The yo-yo in exercise \ref{yoyotrick} from Section \ref{RadianMeasure}.

Recall: The radius of the circle is 28 inches and it completes one revolution in 3 seconds.
\end{problem}

\begin{problem}
    A point on the edge of the hard drive in Exercise \ref{harddrive} from Section \ref{RadianMeasure}.

Recall:  The diameter of the hard disk is 2.5 inches and it spins at 7200 revolutions per minute.
\end{problem}


\begin{problem}\label{motionlast} 
    A passenger on the Big Wheel in Exercise \ref{giantwheelmotion} from Section \ref{RadianMeasure}.

Recall: The diameter is 128 feet and completes 2 revolutions in 2 minutes, 7 seconds.
\end{problem}

\end{question}


\begin{problem}
    Consider the numbers:  $0$, $1$, $2$, $3$, $4$.  Take the square root of each of these numbers, then divide each by $2$. The resulting numbers should look hauntingly familiar. (See the values in the table on \pageref{CosineSineFacts}.) 
\end{problem}

\begin{problem}
    On page \pageref{cosinesineequationsrealnumbers}, we see that the sine and cosine functions of \textit{angles} can be considered functions of \textit{real numbers}.  With help from your classmates, discuss the domains and ranges of  $f(t) = \sin(t)$ and $g(t) = \cos(t)$.  Write your answers using interval notation.
\end{problem}  

\begin{problem}\label{circleofradiusrsinecosinetrans}
      Another way to establish  Theorem \ref{cosinesinecircle} is to use transformations. Re-read the discussion following  Theorem \ref{standardcirclealternate} in Chapter \ref{TheConicSections}  and transform  the Unit Circle, $x^2+y^2 = 1$, to $x^2 + y^2 = r^2$ using horizontal and vertical stretches.  Show if the coordinates on the Unit Circle are $(\cos(\theta), \sin(\theta))$, then the corresponding coordinates on $x^2+y^2 = r^2$ are $(r \cos(\theta), r \sin(\theta))$.
\end{problem}


\begin{problem}
    In the scenario of Equation \ref{equationsforcircularmotion}, we assumed that at $t=0$, the object was at the point $(r,0)$.  If this is not the case,  we can adjust the equations of motion by introducing a `time delay.'   If $t_{0} > 0$ is the first time the object passes through the point $(r,0)$, show, with the help of your classmates, the equations of motion are $x = r \cos(\omega (t - t_{0}))$ and $y = r \sin(\omega (t-t_{0}))$.
\end{problem}


\end{document}
