\documentclass{ximera}

\begin{document}
	\author{Stitz-Zeager}
	\xmtitle{Exercises for Intro Limits}{}

\mfpicnumber{1} \opengraphsfile{ExercisesforIntroLimits} % mfpic settings added 


\label{ExercisesforIntroLimits}

\begin{problem}

Consider the complete graph of the function $f$ below.  Use the graph to find the indicated values.


If a limit fails to exist, state that is the case  or use the symbols `$\infty$' or `$-\infty$' appropriately.
 
 
 \textbf{NOTE:}  The graph has a vertical asymptote $x=3$.  

\begin{center}

\begin{tikzpicture}
\begin{axis}[
    width=5in, height=5in,
    xmin=-10, xmax=10, ymin=-10, ymax=10,
    axis lines=middle,
    axis line style={-{Latex[length=8pt,width=6pt]}},
    tick style={line width=1pt},
    major tick length=5pt,
    xlabel={$x$}, ylabel={$y$},
    xlabel style={anchor=west},
    ylabel style={anchor=south},
    xtick={-8,-6,-4,-2, 0, 2, 4, 6, 8},
    ytick={-8, -6,-4,-2, 0, 2, 4, 6, 8},
    tick label style={font=\tiny},
    label style={font=\scriptsize},
    samples=100,
]
% Parametric Graph

\draw[<-, thick, blue, smooth, samples=100, domain=-10:-3] 
      plot ({\x}, {\x + 6});

\draw[-, thick, blue, smooth, samples=100, domain=-3:3] 
        plot ({\x}, {9 - (\x)^2});

\draw[<->, thick, blue, smooth, samples=100, domain=3.25:10] 
          plot ({\x}, {-2/(\x-3)});
\node at (1.75, -5) {$x = 3$};
\draw[dashed] (3, -9) -- (3, 0);
\draw[dashed] (3, 0) -- (3, 9);
\node at (-3,-1) {$(-3,0)$};
\fill[blue] (-3,0) circle (2pt);
\node at (-1,1) {$(0,1)$};
\fill[blue] (0,1) circle (2pt);
\node at (-3.5,3.75) {$(-3,3)$};
\draw[blue, thick] (-3,3) circle (2pt);
\node at (-1.5,9) {$(0,9)$};
\fill[white, opacity = 1.0] (-3,3)  circle (1.7pt);
\draw[blue, thick] (0,9) circle (2pt);
\node at (4,0.5) {$(3,0)$};
\fill[white, opacity = 1.0] (0,9)  circle (1.7pt);
\draw[blue, thick] (3,0) circle (2pt);
\fill[white, opacity = 1.0] (3,0)  circle (1.7pt);
\end{axis}
\end{tikzpicture}

\end{center}

\begin{enumerate}

\item $\lim_{x \rightarrow -3^{-}} f(x)$

\begin{solution}

$\lim_{x \rightarrow -3^{-}} f(x) = 3$

\end{solution}

\item $\lim_{x \rightarrow -3^{+}} f(x) = \answer{0}$

\item $\lim_{x \rightarrow -3} f(x)$

\begin{solution}

$\lim_{x \rightarrow -3} f(x)$ does not exist. (d.n.e.)

\end{solution}

\item $f(-3) = \answer{0}$

\item $\lim_{x \rightarrow 0} f(x)$

\begin{solution}

$\lim_{x \rightarrow 0} f(x) = 9$

\end{solution}

\item  $f(0) = \answer{1}$

\item $\lim_{x \rightarrow 3^{-}} f(x) = \answer{0}$

\item $\lim_{x \rightarrow 3^{+}} f(x)$

\begin{solution}

$\lim_{x \rightarrow 3^{+}} f(x) = \answer{-\infty}$

\end{solution}

\end{enumerate}

\end{problem}


\begin{problem} \label{twosidedonesidedlimitexistexercise}  

\begin{enumerate}  \item  Explain why if $\lim_{x \rightarrow a} f(x)$ exist, then so do $\lim_{x \rightarrow a^{-}} f(x)$ and $\lim_{x \rightarrow a^{+}} f(x)$.

\begin{solution}

 If $\lim_{x \rightarrow a} f(x)$ exists, say $\lim_{x \rightarrow a} f(x) = L$ then the $f(x)$ values approach  $L$  as $x \rightarrow a$ from both directions.  Hence both one-sided limits exist.  In particular,  $\lim_{x \rightarrow a^{-}} f(x) = L$ and $\lim_{x \rightarrow a^{+}} f(x) = L$.

 \end{solution}

\item   Find an instance\footnote{There are a few examples to be found in Example \ref{limitfromgraphex} \ldots}  where   $\lim_{x \rightarrow a^{-}} f(x)$ and $\lim_{x \rightarrow a^{+}} f(x)$ both exist but $\lim_{x \rightarrow a} f(x)$ does not.

\begin{solution}
In Example \ref{limitfromgraphex},  $\lim_{x \rightarrow -1^{-}} f(x)  = 0$ and $\lim_{x \rightarrow -1^{+}} f(x) = 4$ both exist but $\lim_{x \rightarrow -1} f(x)$ does not because the two one-sided limits are not equal.
\end{solution}

\end{enumerate}

\end{problem}

\begin{problem} Consider the complete graph of the function $g$ below.  Use the graph to find the indicated values.

If a limit fails to exist, state that is the case  or use the symbols `$\infty$' or `$-\infty$' appropriately.
  
 \textbf{NOTE:}  The graph has a vertical asymptote $x=-2$ and a horizontal asymptote $y = 1$.  


\begin{center}

\begin{tikzpicture}
\begin{axis}[
    width=5in, height=5in,
    xmin=-10, xmax=10, ymin=-10, ymax=10,
    axis lines=middle,
    axis line style={-{Latex[length=8pt,width=6pt]}},
    tick style={line width=1pt},
    major tick length=5pt,
    xlabel={$x$}, ylabel={$y$},
    xlabel style={anchor=west},
    ylabel style={anchor=south},
    xtick={-8,-6,-4,-2, 0, 2, 4, 6, 8},
    ytick={-8, -6,-4,-2, 0, 2, 4, 6, 8},
    tick label style={font=\tiny},
    label style={font=\scriptsize},
    samples=100,
]
% Parametric Graph

\draw[<->, thick, blue, smooth, samples=100, domain=-9:-2.20] 
      plot ({\x}, {(\x+4)/(\x+2)});
\draw[<-, thick, blue, smooth, samples=100, domain=-1.75:2] 
      plot ({\x}, {(\x+4)/(\x+2)});
\draw[->, thick, blue, smooth, samples=100, domain=2:9] 
      plot ({\x}, {(2-\x)/4});

\draw[dashed] (-2, -9) -- (-2, 0);
\draw[dashed] (-2, 0) -- (-2, 9);
\node at (-3.25, 3) {$x = -2$};

\draw[dashed] (-9, 1) -- (9, 1);
\node at (-5, 1.5) {$y = 1$};

\node at (1,2.5) {$(0,2)$};
\draw[blue, thick] (0,2) circle (2pt);
\fill[white, opacity = 1.0] (0,2) circle (1.7pt);
\node at (3.25,1.5) {$(2,1.5)$};
\fill[blue] (2,1.5) circle (2pt);
\node at (2,-1) {$(2,0)$};
\fill[white, opacity = 1.0] (0,9)  circle (1.7pt);
\draw[blue, thick] (2,0) circle (2pt);
\fill[white, opacity = 1.0] (2,0)  circle (1.7pt);
\end{axis}
\end{tikzpicture}

\end{center}

\begin{enumerate}

\item $\lim_{x \rightarrow -\infty} g(x)$

\begin{solution}
$\lim_{x \rightarrow -\infty} g(x) = 1$
\end{solution}

\item $\lim_{x \rightarrow -2^{-}} g(x) = \answer{-\infty}$

\item $\lim_{x \rightarrow -2^{+}} g(x)$

\begin{solution}
$\lim_{x \rightarrow -2^{+}} g(x) = \infty$
\end{solution}

\item  $\lim_{x \rightarrow \infty} g(x) = \answer{ -\infty}$

\item $\lim_{x \rightarrow 0} g(x)$

\begin{solution}
$\lim_{x \rightarrow 0} g(x) = 2$
\end{solution}

\item $\lim_{x \rightarrow 2^{-}} g(x) = \answer{1.5}$

\item  $g(2)$

\begin{solution}
$g(2) = 1.5$
\end{solution}

\item $\lim_{x \rightarrow 2^{+}} g(x) = \answer{0}$

\end{enumerate}

\end{problem}

\begin{question}
For Exercises \ref{factorcancelfirst} - \ref{factorcancellast}, find the limit analytically using Exercise \ref{rationallimit} as a guide.  If a limit fails to exist, state that is the case  or use the symbols `$\infty$' or `$-\infty$' appropriately.
 
\begin{problem} \label{factorcancelfirst}  
$\lim_{x \rightarrow 2} \frac{2x^2+x-3}{x^2-1}$

\begin{solution}
$\lim_{x \rightarrow 2} \frac{2x^2+x-3}{x^2-1} = \frac{7}{3}$
\end{solution}
\end{problem}
  
\begin{problem}
$\lim_{x \rightarrow 1} \frac{2x^2+x-3}{x^2-1} = \answer{\frac{5}{2}}$
\end{problem}

\begin{problem}
$\lim_{x \rightarrow -1} \frac{2x^2+x-3}{x^2-1}$

\begin{solution}
$\lim_{x \rightarrow -1} \frac{2x^2+x-3}{x^2-1}$ does not exist (d.n.e.)
\end{solution}
\end{problem}


\begin{problem}
$\lim_{x \rightarrow -1^{+}} \frac{2x^2+x-3}{x^2-1} = \answer{\infty}$
\end{problem}


\begin{problem}
$\lim_{x \rightarrow 3} \frac{x^2-3x}{x^2-x-6}$

\begin{solution}
$\lim_{x \rightarrow 3} \frac{x^2-3x}{x^2-x-6} = \frac{3}{5}$
\end{solution}
\end{problem}

\begin{problem} \label{factorcancellast}
$\lim_{x \rightarrow 3} \frac{x^2-6x}{x^2-6x+9} = \answer{-\infty}$
\end{problem}

\end{question}

\begin{question}
For Exercises \ref{abscancelfirst} - \ref{abscancellast}, use the piecewise definition of absolute value, Definition \ref{absolutevaluepiecewise} to help you find the limit analytically.    If a limit fails to exist, state that is the case  or use the symbols `$\infty$' or `$-\infty$' appropriately.

\begin{problem} \label{abscancelfirst} 
$\lim_{x \rightarrow 3^{-}} \frac{|3x - x^2| }{x-3} = \answer{-3}$ 

\end{problem}

\begin{problem} \label{abscancellast}  
$\lim_{x \rightarrow 2^{+}} \frac{|6-3x|}{x^2-4x+4}$

\begin{solution}
$\lim_{x \rightarrow 2^{+}} \frac{|6-3x|}{x^2-4x+4} = \infty$
\end{solution}
\end{problem}

\end{question}

\begin{question}
For Exercises \ref{complexcancelfirst} - \ref{complexcancellast}, simplify the complex fraction in order to help you find the limit analytically.\footnote{A review of Section \ref{AppRatExpEqus} may be in order.}  If a limit fails to exist, state that is the case  or use the symbols `$\infty$' or `$-\infty$' appropriately.

\begin{problem}\label{complexcancelfirst}  
$\lim_{x \rightarrow 1} \frac{\frac{x}{x-2} +1}{x-1}$

\begin{solution}
$\lim_{x \rightarrow 1} \frac{\frac{x}{x-2} +1}{x-1} = -2$
\end{solution}
\end{problem}

\begin{problem}
$\lim_{x \rightarrow 2} \frac{\frac{2x}{x+2}  -1}{x-2} = \answer{\frac{1}{4}}$ 
\end{problem}
     
\begin{problem}\label{complexcancellast} 
$\lim_{h \rightarrow 0} \frac{\frac{1}{2(x+h) - 1}  - \frac{1}{2x-1} }{h}$ 

\begin{solution}
$\lim_{h \rightarrow 0} \frac{\frac{1}{2(x+h) - 1}  - \frac{1}{2x-1} }{h} = -\frac{2}{(2x-1)^2}$
\end{solution}
\end{problem}

\end{question}

\begin{question}
For Exercises \ref{radicalcancelfirst} - \ref{radicalcancellast}, rationalize the numerator of the fraction in order to help you find the limit analytically.\footnote{A review of Section \ref{rationalizingdenomandnumer} may be in order.} If a limit fails to exist, state that is the case  or use the symbols `$\infty$' or `$-\infty$' appropriately.


\begin{problem} \label{radicalcancelfirst}  
$\lim_{x \rightarrow -1} \frac{\sqrt{x+5} - 2 }{x+1}$

\begin{solution}
$\lim_{x \rightarrow -1} \frac{\sqrt{x+5} - 2 }{x+1}  = \frac{1}{4}$
\end{solution}
\end{problem}
   
\begin{problem} 
$\lim_{h \rightarrow 0} \frac{\sqrt{4+h} - 2}{h} = \answer{\frac{1}{4}}$
\end{problem}

\begin{problem}\label{radicalcancellast} 
$\lim_{h \rightarrow 0} \frac{\sqrt{2x+2h-1} - \sqrt{2x-1} }{h}$

\begin{solution}
$\lim_{h \rightarrow 0} \frac{\sqrt{2x+2h-1} - \sqrt{2x-1} }{h}  = \frac{1}{\sqrt{2x-1}}$
\end{solution}
\end{problem}

\end{question}

\begin{question}
 In Exercises \ref{limitatinfinityfirst} - \ref{limitatinfinitylast},  find the limit analytically.  Use the symbols `$\infty$' and `$-\infty$' as appropriate.

\begin{problem} \label{limitatinfinityfirst}   
$\lim_{x \rightarrow \infty} \frac{3x-4}{2x+1}$

\begin{solution}
$\lim_{x \rightarrow \infty} \frac{3x-4}{2x+1} = \frac{3}{2}$
\end{solution}
\end{problem}

\begin{problem}
$\lim_{x \rightarrow -\infty}  \frac{1-2x}{x-5} = \answer{-2}$
\end{problem}

\begin{problem}
$\lim_{x \rightarrow -\infty}\frac{2x-1}{x^2+4}$

\begin{solution}
$\lim_{x \rightarrow -\infty}\frac{2x-1}{x^2+4} = 0$
\end{solution}
\end{problem}

\begin{problem}
$\lim_{x \rightarrow \infty}\frac{\sqrt{4x^2+x-1}}{1-x} = \answer{-2}$
\end{problem}

\begin{problem}
$\lim_{x \rightarrow -\infty}\frac{\sqrt{4x^2+x-1}}{1-x}$

\begin{solution}
$\lim_{x \rightarrow -\infty}\frac{\sqrt{4x^2+x-1}}{1-x}  = 2$
\end{solution}
\end{problem}

\begin{problem} \label{limitatinfinitylast}  
$\lim_{x \rightarrow \infty} \frac{x^2+2x+3}{4-x} = \answer{-\infty}$.
\end{problem}

\end{question}

\begin{problem} 
Let $f(x) = \frac{x}{\lfloor x \rfloor}$  where `$\lfloor x \rfloor$'  is the greatest integer (or floor) function.\footnote{See Example \ref{greatestintegerdefn} in Section \ref{ConstantandLinearFunctions} for review, if needed.} 



Fill in the blanks below to help you analyze  $\lim_{x \rightarrow 0} f(x)$.


\begin{enumerate}

\item  If $-1 < x < 0$, then $\lfloor x \rfloor = \answer{-1}$. So we can rewrite $f(x) = \frac{x}{\lfloor x \rfloor} = \answer{-x}$.

\item  Using part (a), we can find  $\lim_{x \rightarrow 0^{-}} f(x) =  \lim_{x \rightarrow 0^{-}} (\answer{-x}) = \answer{0}$.

\item  If $0 < x < 1$, then $\lfloor x \rfloor =\answer{0}$, hence $f(x) = \frac{x}{\lfloor x \rfloor}$ does not exist as $x \rightarrow 0^{+}$.

\item  Putting parts (b) and (c) together, we have that   $\lim_{x \rightarrow 0} f(x)$ does not exist.

\item  Graph $f(x) = \frac{x}{\lfloor x \rfloor}$ on desmos near $x = 0$ to confirm your answers.

\end{enumerate}

\end{problem}

\begin{problem}
With help from your classmates, sketch the graph of a function which satisfies all of the following criteria:

\begin{itemize}

\item $\lim_{x \rightarrow -\infty}  f(x) = \infty$

\item $\lim_{x \rightarrow 4^{-}} f(x) = 6$

\item $\lim_{x \rightarrow 4^{+}} f(x) = - \infty$

\item $\lim_{x \rightarrow \infty}  f(x) =0$

\end{itemize}

\end{problem}

\begin{problem}
With help from your classmates, sketch  the graph of a function $f$  which satisfies all of the following criteria:

\begin{itemize}

\item $\lim_{x \rightarrow -\infty} f(x) = 2$

\item $\lim_{x \rightarrow 0^{-}} f(x) = \infty$

\item $\lim_{x \rightarrow 0^{+}} f(x) = -\infty$

\item $\lim_{x \rightarrow 2^{-}} f(x) = 3$

\item $\lim_{x \rightarrow 2^{+}} f(x) =0$

\item $\lim_{x \rightarrow \infty} f(x) = -\infty$

\end{itemize}

\end{problem}


\begin{problem}\label{onesidedcontinuityexercise}  
A function is said to be \index{continuous from the left}\index{continuous ! from the left}\index{one-sided continuity}\index{continuity ! one sided}\textbf{continuous from the left} at $x=a$ if $\lim_{x \rightarrow a^{-}} f(x) = f(a)$.  Likewise, a function is said to be \index{continuous from the right}\index{continuous ! from the right}\index{one-sided continuity}\index{continuity ! one sided}\textbf{continuous from the right} at $x=a$ if $\lim_{x \rightarrow a^{+}} f(x) = f(a)$.

\begin{enumerate}

\item   Explain why $r(x) = \sqrt{5-x}$ is not continuous at $x = 5$.  Is $r$ continuous from the left at $x=5$?  From the right?  Explain.

\begin{solution}
The function $r(x) = \sqrt{5-x}$ is not continuous at $x = 5$ since $r$ is undefined if $x>5$, so $\lim_{x \rightarrow 5^{+}} r(x)$ does not exist. However, $r$ is continuous from the left at $x = 5$ since 

\[\lim_{x \rightarrow 5^{-}} r(x) =  \lim_{x \rightarrow 5^{-}} \sqrt{5-x} = 0 = \sqrt{5-5} = r(5).\]

\end{solution}

\item  Explain why the floor function\footnote{See Example \ref{greatestintegerdefn} in Section \ref{ConstantandLinearFunctions} for review, if needed.} $F(x) = \lfloor x \rfloor$ is not continuous at $x = 117$.   Is $F$ continuous from the left at $x=117$?  From the right?  Explain.

\begin{solution}
The function  $F(x) = \lfloor x \rfloor$ is not continuous at $x = 117$ since $\lim_{x \rightarrow 117^{-}} \lfloor x \rfloor = 116$ but $\lim_{x \rightarrow 117^{+}} \lfloor x \rfloor = 117$.  However, $F$ is continuous from the right at $x = 117$ since $\lim_{x \rightarrow 117^{+}} \lfloor x \rfloor = 117 = \lfloor 117 \rfloor = F(117)$. 
\end{solution}

\item  If a function $f$ is continuous at $x = a$, explain why $f$ is continuous from both the left and the right at $x = a$.  Is the converse true?  That is, if $f$ is continuous from both the left and the right at $x = a$, is $f$ continuous at $x=a$?  

\begin{solution}
If $f$ is continuous at $x=a$, then $\lim_{x \rightarrow a} f(x) = f(a)$.  This means  $\lim_{x \rightarrow a^{-}} f(x) = f(a)$ and  $\lim_{x \rightarrow a^{+}} f(x) = f(a)$, so $f$ is continuous from both directions at $x=a$. The converse is also true since if $\lim_{x \rightarrow a^{-}} f(x) = f(a)$ and  $\lim_{x \rightarrow a^{+}} f(x) = f(a)$, then $\lim_{x \rightarrow a} f(x) = f(a)$.

\end{solution}

\item  Compare and contrast your answers in this exercise to those in Exercise \ref{twosidedonesidedlimitexistexercise}.

\begin{solution}
The difference between the scenario here and that in Exercise \ref{twosidedonesidedlimitexistexercise} is that here, we know what each of the one-sided limits are: $f(a)$. They can't be different numbers like they could be in  Example \ref{limitfromgraphex}.
\end{solution}
\end{enumerate}

\end{problem}

\begin{problem}
Consider the table of values below:

\[ \begin{array}{|r|c|}  \hline
x & f(x)  \\    \hline 
-0.001  &  1.9 \\   \hline 
-0.0001 & 1.99 \\   \hline
-0.00001 & 1.999\\   \hline
-0.000001 & 1.9999 \\  \hline
0.000001 & -10000\\   \hline
0.00001 & -1000 \\  \hline
0.0001 & - 100 \\   \hline
0.001 & -10 \\  \hline
\end{array} \]

It turns out that $\lim_{x \rightarrow 0} f(x) = 117$.  How is this possible assuming the data in the table is correct?

\begin{solution}
We are told nothing of the function on the interval $(-0.000001, 0.000001 )$ so the function has plenty of opportunities to approach $117$. 
\end{solution}

\end{problem}

\begin{problem} 
In this exercise, we use Definition \ref{infinitelimitsatinfinity} to prove $\lim_{x \rightarrow \infty} x^3 = \infty$ and $\lim_{x \rightarrow -\infty} x^3 = -\infty$:


\begin{enumerate}

\item  Solve the following inequalities:



\begin{itemize}

\item  $x^3 > 1000$

\begin{solution}
$x^3 > 1000$ for $x > 10$
\end{solution}

\item  $x^3 > 100000$

\begin{solution}
$x^3 > 1000000$ for $x > 100$
\end{solution}

\item  $x^3 > 10^{99}$

\begin{solution}
$x^3 > 10^{99}$ for $x > 10^{33}$
\end{solution}


\item  $x^3 > N$

\begin{solution}
 $x^3 > N$ for $x > \sqrt[3]{N}$
\end{solution}

\end{itemize}

\item  Show that for $N > 0$, if $x > \sqrt[3]{N}$, then $x^3 > N$.  Write a sentence (or two!) which uses your work and Definition \ref{infinitelimitsatinfinity} to prove $\lim_{x \rightarrow \infty} x^3 = \infty$.

\begin{solution}
If $x > \sqrt[3]{N}$, then $x^3 > \left(\sqrt[3]{N} \right)^3 = N$.  Per  Definition \ref{infinitelimitsatinfinity}, given $N>0$, choose $M = \sqrt[3]{N}$.  If $x > M$, then $x^3 > M^3 = N$. Hence,  $\lim_{x \rightarrow \infty} x^3 = \infty$.
\end{solution}

\item  Repeat a similar argument to prove $\lim_{x \rightarrow -\infty} x^3 = -\infty$.

\begin{solution}
Given $N<0$, choose $M = \sqrt[3]{N}$.  If $x< M$, then $x^3 < M^3 = \left(\sqrt[3]{N}\right)^3 = N$.  Hence,   $\lim_{x \rightarrow -\infty} x^3 = -\infty$.
\end{solution}
\end{enumerate}
 
\end{problem}

\end{document}


