\documentclass{ximera}

\begin{document}
	\author{Stitz-Zeager}
	\xmtitle{Exercises for Function Composition}{}

\mfpicnumber{1} \opengraphsfile{ExercisesforFunctionComposition} % mfpic settings added 


\label{ExercisesforFunctionComposition}


\begin{question}
    
In Exercises \ref{funccompeval1first} - \ref{funccompeval1last}, use the given pair of functions to find the following values if they exist.




\begin{itemize}

\item  $(g\circ f)(0)$

\item  $(f\circ g)(-1)$

\item  $(f \circ f)(2)$

\end{itemize}



\begin{itemize}

\item  $(g\circ f)(-3)$

\item  $(f\circ g)\left(\frac{1}{2}\right)$

\item  $(f \circ f)(-2)$

\end{itemize}



\begin{problem}\label{funccompeval1first}
    $f(x) = x^2$, $g(t) = 2t+1$ 
\end{problem}

\begin{problem}

$f(x) = 4-x$, $g(t) = 1-t^2$
    
\end{problem}

\begin{problem}
    $f(x) = 4-3x$, $g(t) = |t|$
\end{problem}

\begin{problem}
    $f(x) = |x-1|$, $g(t) = t^2-5$
\end{problem}

\begin{problem}
$f(x) = 4x+5$, $g(t) = \sqrt{t}$
\end{problem}

\begin{problem}
$f(x) = \sqrt{3-x}$, $g(t) = t^2+1$
\end{problem}
  
\begin{problem}
$f(x) = 6-x-x^2$, $g(t) = t\sqrt{t+10}$
\end{problem}  

\begin{problem}
$f(x) = \sqrt[3]{x+1}$, $g(t) = 4t^2-t$
\end{problem}  

\begin{problem}
$f(x) = \dfrac{3}{1-x}$, $g(t) = \dfrac{4t}{t^2+1}$
\end{problem} 


\begin{problem}
 $f(x) = \dfrac{x}{x+5}$, $g(t) = \dfrac{2}{7-t^2}$
\end{problem}

\begin{problem}
$f(x) = \dfrac{2x}{5-x^2}$, $g(t) = \sqrt{4t+1}$
\end{problem}

\begin{problem}\label{funccompeval1last}
$f(x) =\sqrt{2x+5}$, $g(t) = \dfrac{10t}{t^2+1}$ 
\end{problem}

\end{question}

\begin{question}
In Exercises \ref{funccompexp1first} - \ref{funccompexp1last}, use the given pair of functions to find and simplify expressions for the following functions and state the domain of each using interval notation.

\begin{multicols}{3}

\begin{itemize}

\item  $(g \circ f)(x)$

\item  $(f \circ g)(t)$

\item  $(f \circ f)(x)$


\end{itemize}

\end{multicols}

\begin{problem}\label{funccompexp1first}
$f(x) = 2x+3$, $g(t) = t^2-9$  
\end{problem}

\begin{problem}
$f(x) = x^2 -x+1$, $g(t) = 3t-5$ 
\end{problem}
 
\begin{problem}
$f(x) = x^2-4$, $g(t) = |t|$
\end{problem}
 
\begin{problem}
$f(x) = 3x-5$, $g(t) = \sqrt{t}$ 
\end{problem}

\begin{problem}
$f(x) = |x+1|$, $g(t) = \sqrt{t}$  
\end{problem}

\begin{problem}
$f(x) = 3-x^2$, $g(t) = \sqrt{t+1}$  
\end{problem}

\begin{problem}
$f(x) = |x|$, $g(t) = \sqrt{4-t}$  
\end{problem}

\begin{problem}
$f(x) = x^2-x-1$, $g(t) = \sqrt{t-5}$  
\end{problem}

\begin{problem}
$f(x) = 3x-1$, $g(t) = \dfrac{1}{t+3}$ 
\end{problem}

\begin{problem}
$f(x) = \dfrac{3x}{x-1}$, $g(t) =\dfrac{t}{t-3}$
\end{problem}

\begin{problem}
$f(x) = \dfrac{x}{2x+1}$, $g(t) = \dfrac{2t+1}{t}$
\end{problem}

\begin{problem}\label{funccompexp1last}
$f(x) =  \dfrac{2x}{x^2-4}$, $g(t) =\sqrt{1-t}$
\end{problem}

\end{question}

\begin{question}
In Exercises \ref{threefunccompfirst} - \ref{threefunccomplast}, use $f(x) = -2x$, $g(t) = \sqrt{t}$ and $h(s) = |s|$ to find and simplify expressions for the following functions and state the domain of each using interval notation.

\begin{problem}\label{threefunccompfirst}
$(h\circ g \circ f)(x)$ 
\end{problem}

\begin{problem}
$(h\circ f \circ g)(t)$
\end{problem}
 
\begin{problem}
$(g\circ f \circ h)(s)$
\end{problem}

\begin{problem}
$(g\circ h \circ f)(x)$
\end{problem}  

\begin{problem}
$(f\circ h \circ g)(t)$
\end{problem} 

\begin{problem}\label{threefunccomplast}
$(f\circ g \circ h)(s)$ 
\end{problem} 

\end{question}

\begin{question}
In Exercises \ref{pointcompexfirst} - \ref{pointcompexlast}, let $f$ be the function defined by \[f = \{(-3, 4), (-2, 2), (-1, 0), (0, 1), (1, 3), (2, 4), (3, -1)\}\] and let $g$ be the function defined by \[g = \{(-3, -2), (-2, 0), (-1, -4), (0, 0), (1, -3), (2, 1), (3, 2)\}.\]  Find the following, if it exists.

\begin{problem}\label{pointcompexfirst}
$(f \circ g)(3)$
\end{problem}

\begin{problem}
$f(g(-1))$
\end{problem}

\begin{problem}
$(f \circ f)(0)$
\end{problem}

\begin{problem}
$(f \circ g)(-3)$
\end{problem}

\begin{problem}
$(g \circ f)(3)$
\end{problem}

\begin{problem}
$g(f(-3))$
\end{problem}

\begin{problem}
$(g \circ g)(-2)$
\end{problem}

\begin{problem}
$(g \circ f)(-2)$
\end{problem}

\begin{problem}
$g(f(g(0)))$
\end{problem}

\begin{problem}
$f(f(f(-1)))$
\end{problem}

\begin{problem}
$f(f(f(f(f(1)))))$
\end{problem}

\begin{problem}
$\underbrace{(g \circ g \circ \cdots \circ g)}_{\mbox{$n$ times}}(0)$ 
\end{problem}

\begin{problem}\label{pointcompexlast}
Find the domain and range of $f \circ g$ and $g \circ f$. 
\end{problem}  

\end{question}

\begin{question}
    

In Exercises \ref{twofuncgraphcompfirst} - \ref{twofuncgraphcomplast}, use the graphs of $y=f(x)$ and $y=g(x)$ below to find the following if it exists.

\begin{image}
\begin{tikzpicture}[font=\LARGE]

% -------- LEFT GRAPH: y = f(x) --------
\begin{axis}[
    at={(0,0)},
    anchor=origin,
    domain=-1:5,
    xmin=-1, xmax=5,
    ymin=-1, ymax=5,
    width=3in,
    height=3in,
    xtick={1,2,3,4},
    ytick={1,2,3,4},
    tick style={line width=1.2pt},
    major tick length=6pt,
    axis x line=center,
    axis y line=center,
    x axis line style={-{Latex[length=10pt,width=8pt]}, line width=1.2pt},
    y axis line style={-{Latex[length=10pt,width=8pt]}, line width=1.2pt},
    xlabel={$x$},
    ylabel={$y$},
    every axis x label/.style={
        at={(current axis.right of origin)},
        anchor=west
    }
]

% Polyline
\addplot[thick, line width=3pt] coordinates {
    (0,1) (1,1) (2,3) (2.5,4) (3,3) (4,0)
};

% Points
\addplot[only marks, mark=*, mark size=4pt] coordinates {
    (0,1) (1,1) (2,3) (2.5,4) (3,3) (4,0)
};

% Labels (carefully placed)
%\node[anchor=east, xshift=-15pt] at (axis cs:0,1) {\Large $(0,1)$};
\node[anchor=north, yshift=-6pt] at (axis cs:1,1) {\Large $(1,1)$};
\node[anchor=east, xshift=-6pt] at (axis cs:2,3) {\Large $(2,3)$};
\node[anchor=south, yshift=8pt] at (axis cs:2.5,4) {\Large $(2.5,4)$};
\node[anchor=west, xshift=6pt] at (axis cs:3,3) {\Large $(3,3)$};
\node[anchor=north, yshift=-16pt] at (axis cs:4,0) {\Large $(4,0)$};

% Caption
\node at (axis cs:2,-0.8) {\Large $y=f(x)$};

\end{axis}

% -------- RIGHT GRAPH: y = g(x) --------
\begin{axis}[
    at={(4.5in,0)},
    anchor=origin,
    domain=-1:5,
    xmin=-1, xmax=5,
    ymin=-1, ymax=5,
    width=3in,
    height=3in,
    xtick={1,2,3,4},
    ytick={1,2,3,4},
    tick style={line width=1.2pt},
    major tick length=6pt,
    axis x line=center,
    axis y line=center,
    x axis line style={-{Latex[length=10pt,width=8pt]}, line width=1.2pt},
    y axis line style={-{Latex[length=10pt,width=8pt]}, line width=1.2pt},
    xlabel={$x$},
    ylabel={$y$},
    every axis x label/.style={
        at={(current axis.right of origin)},
        anchor=west
    }
]

% Polyline
\addplot[thick, line width=3pt] coordinates {
    (0,0) (1,3) (2,3) (3,0)
};

% Points
\addplot[only marks, mark=*, mark size=4pt] coordinates {
    (0,0) (1,3) (2,3) (3,0)
};

% Labels (kept off axes & lines)
%\node[anchor=north, yshift=-6pt] at (axis cs:0,0) {\Large $(0,0)$};
\node[anchor=south, xshift=-4pt, yshift=6pt] at (axis cs:1,3) {\Large $(1,3)$};
\node[anchor=south, xshift=4pt, yshift=6pt] at (axis cs:2,3) {\Large $(2,3)$};
\node[anchor=north, xshift=15pt, yshift=-15pt] at (axis cs:3,0) {\Large $(3,0)$};

% Caption
\node at (axis cs:2,-0.8) {\Large $y=g(x)$};

\end{axis}

\end{tikzpicture}
\end{image}



\begin{problem}\label{twofuncgraphcompfirst}
$(g\circ f)(1)$ 
\end{problem}

\begin{problem}
$(f \circ g)(3)$
\end{problem}

\begin{problem}
$(g\circ f)(2)$
\end{problem}

\begin{problem}
$(f\circ g)(0)$
\end{problem}

\begin{problem}
$(f\circ f)(4)$
\end{problem}

\begin{problem}
$(g \circ g)(1)$
\end{problem}

\begin{problem}\label{twofuncgraphcomplast}
Find the domain and range of $f \circ g$ and $g \circ f$.
\end{problem}

\end{question}

\begin{question}
In Exercises \ref{breakdowncompexfirst} - \ref{breakdownxomexlast},  write the given function as a composition of two or more non-identity functions.  (There are several correct answers, so check your answer using function composition.)

\begin{problem}\label{breakdowncompexfirst}
$p(x) = (2x+3)^3$ 
\end{problem}

\begin{problem}
$P(x) = \left(x^2-x+1\right)^5$
\end{problem} 

\begin{problem}
$h(t) = \sqrt{2t-1}$
\end{problem}  

\begin{problem}
$H(t) = |7-3t|$
\end{problem}

\begin{problem}
$r(s) = \dfrac{2}{5s+1}$
\end{problem} 

\begin{problem}
$R(s) = \dfrac{7}{s^2-1}$
\end{problem}

\begin{problem}
$q(z) = \dfrac{|z|+1}{|z|-1}$
\end{problem} 

\begin{problem}
$Q(z) = \dfrac{2z^3+1}{z^3-1}$
\end{problem} 

\begin{problem}
$v(x) = \dfrac{2x+1}{3-4x}$
\end{problem} 

\begin{problem}\label{breakdownxomexlast}
$w(x) = \dfrac{x^2}{x^4+1}$ 
\end{problem}  

\end{question}

\begin{problem}
Write the function $F(x) = \sqrt{\dfrac{x^{3} + 6}{x^{3} - 9}}$ as a composition of three or more non-identity functions.
\end{problem}

\begin{problem}
Let $g(x) = -x, \, h(x) = x + 2, \, j(x) = 3x$ and $k(x) = x - 4$.  In what order must these functions be composed with $f(x) = \sqrt{x}$ to create $F(x) = 3\sqrt{-x + 2} - 4$?
\end{problem}

\begin{problem}
What linear functions could be used to transform $f(x) = x^{3}$ into $F(x) = -\frac{1}{2}(2x - 7)^{3} + 1$?  What is the proper order of composition?
\end{problem}

\begin{problem}
Let $f(x) = 3x+1$ and let $g(x) =    \begin{cases}  2x-1 &  \text{if $x \leq 3$} \\   4-x & \text{if $x > 3$} \\  \end{cases}$.  Find expressions for $(f \circ g)(x)$ and $(g \circ f)(x)$.

\end{problem}

\begin{problem}
The volume $V$ of a cube is a function of its side length $x$.  Let's assume that $x = t + 1$ is also a function of time $t$, where $x$ is measured in inches and $t$ is measured in minutes.  Find a formula for $V$ as a function of $t$.
\end{problem}

\begin{problem}
Suppose a local vendor charges $\$2$ per hot dog and that the number of hot dogs sold per hour $x$ is given by $x(t) = -4t^2+20t+92$, where $t$ is the number of hours since $10$ AM, $0 \leq t \leq 4$.

\begin{enumerate}

\item  Find an expression for the revenue per hour $R$ as a function of $x$.
\item  Find and simplify $\left(R \circ x\right)(t)$.  What does this represent?
\item  What is the revenue per hour at noon?
\item Using Example \ref{surfaceareaex2chainrule} as a guide, verify $\frac{\Delta[R(x)]}{\Delta x} \cdot \frac{\Delta[x(t)]}{\Delta t} = \frac{\Delta[R(t)]}{\Delta t}$.

\end{enumerate}
\end{problem}

\begin{problem}
The book in Example \ref{dragforceex} plunges into a lake and generates a circular wave pattern.  If the waves are tracked as traveling at a constant $0.5$ meters per second $\left( \frac{\text{m}}{\text{s}}\right)$, use Theorem \ref{relatedratesaroc} to find the rate at which the area of the disturbance is changing with respect to time as the radius changes from $r = 1$ to $r = 1.1$ meters (m).  Be sure to include units on your answer.

\begin{hint}
Recall the area, $A$, enclosed by a circle of radius $r$ is given by $A = \pi \, r^2$.  Here, $\frac{\Delta r}{\Delta t} = 0.5 \, \frac{\text{m}}{\text{s}}$.
\end{hint}
\end{problem}

\begin{problem}
Perfectly fine precalculus textbooks which have no Calculus content are being fed into a shredder at a rate of 2 books per minute in order to make room for precalculus textbooks with Calculus content.  The shredder creates a pile of debris which is in the shape of a right circular cone whose height is twice its width. 

\begin{enumerate}

\item  Assume the volume of the conical pile, $V$,  is given by $V = \frac{1}{3} \, \pi r^2 h$ where $r$ is the radius of the base of the pile and $h$ is the height of the pile.  Given the pile is twice as tall as it is wide, show we can write $V = \frac{4}{3} \, \pi r^3$.

\item  Assuming a typical precalculus textbook is $0.10$ cubic feet $\left( \text{ft}^3 \right)$, use Theorem \ref{relatedratesaroc} to find the rate of change of the radius of the pile with respect to time as the radius changes from $2$ to $2.1$ feet. Be sure to include units on your answer.

\end{enumerate}
\end{problem}

\begin{problem}
Discuss with your classmates how `real-world' processes such as filling out federal income tax forms or computing your final course grade could be viewed as a use of function composition.  Find a process for which composition with itself (iteration) makes sense.
\end{problem}

\end{document}
