\documentclass{ximera}

\begin{document}
	\author{Stitz-Zeager}
	\xmtitle{Exercises for Trigonometric Equations and Inequalities}{}

\mfpicnumber{1} \opengraphsfile{ExercisesforTrigonometricEquationsandInequalities} % mfpic settings added 

\begin{question}
In Exercises \ref{solvebasicfirst} - \ref{solvebasiclast}, find \underline{all} of the exact solutions of the  equation and then list those solutions which are in the interval $[0, 2\pi)$.

\begin{problem} \label{solvebasicfirst}
$\sin \left( 5 \theta \right) = 0$ 
\end{problem}

\begin{problem}
$\cos \left( 3t \right) = \frac{1}{2}$
\end{problem}

\begin{problem}
$\sin \left( -2x \right) = \frac{\sqrt{3}}{2}$
\end{problem}

\begin{problem}
$\tan \left( 6 \theta \right) = 1$
\end{problem}

\begin{problem}
$\csc \left( 4 t \right) = -1$
\end{problem}

\begin{problem}
$\sec \left( 3x \right) = \sqrt{2}$
\end{problem}

\begin{problem}
$\cot \left( 2 \theta \right) = -\frac{\sqrt{3}}{3}$
\end{problem}

\begin{problem}
$\cos \left( 9t  \right) = 9$
\end{problem}

\begin{problem}
$\sin \left( \frac{x}{3} \right) = \frac{\sqrt{2}}{2}$
\end{problem}

\begin{problem}
$\cos \left( \theta+ \frac{5\pi}{6} \right) = 0$
\end{problem}

\begin{problem}
$\sin \left( 2t - \frac{\pi}{3} \right) = -\frac{1}{2}$
\end{problem}

\begin{problem}
$2\cos \left( x + \frac{7\pi}{4} \right) = \sqrt{3}$
\end{problem}

\begin{problem}
$\csc( \theta) = 0$
\end{problem}

\begin{problem}
$\tan \left( 2t - \pi \right) = 1$
\end{problem}

\begin{problem}
$\tan^{2} \left( x \right) = 3$
\end{problem}

\begin{problem}
$\sec^{2} \left( \theta \right) = \frac{4}{3}$
\end{problem}

\begin{problem}
$\cos^{2} \left( t \right) = \frac{1}{2}$
\end{problem}

\begin{problem} \label{solvebasiclast}
$\sin^{2} \left( x \right) = \frac{3}{4}$ 
\end{problem}
\end{question}

\begin{question}
In Exercises \ref{solveidentfirst} - \ref{solveidentlast}, solve the equation, giving the exact solutions which lie in $[0, 2\pi)$

\begin{problem} \label{solveidentfirst}
$\sin \left( \theta \right) = \cos \left( \theta \right)$ 
\end{problem}

\begin{problem}
$\sin \left( 2t \right) = \sin \left( t \right)$
\end{problem}

\begin{problem}
$\sin \left( 2x \right) = \cos \left( x \right)$
\end{problem}

\begin{problem}
$\cos \left( 2\theta \right) = \sin \left( \theta \right)$
\end{problem}

\begin{problem}
$\cos \left( 2t \right) = \cos \left( t \right)$
\end{problem}

\begin{problem}
$\cos(2x) = 2 - 5\cos(x)$
\end{problem}

\begin{problem}
$3\cos(2 \theta ) + \cos(\theta) + 2 = 0$
\end{problem}

\begin{problem}
$\cos(2t) = 5\sin(t) - 2$
\end{problem}

\begin{problem}
 $3\cos(2x) = \sin(x) + 2$
 \end{problem}
 
\begin{problem} 
$2\sec^{2}(\theta) = 3 - \tan(\theta)$
\end{problem}

\begin{problem}
$\tan^{2}(t) = 1-\sec(t)$
\end{problem}


\begin{problem}
$\cot^{2}(x) = 3\csc(x) - 3$
\end{problem}

\begin{problem}
$\sec(\theta) = 2\csc(\theta)$
\end{problem}

\begin{problem}
$\cos(t) \csc(t)\cot(t) = 6-\cot^{2}(t)$
\end{problem}

\begin{problem}
$\sin(2x) = \tan(x)$
\end{problem}

\begin{problem}
$\cot^{4}(\theta) = 4\csc^{2}(\theta) - 7$
\end{problem}

\begin{problem}
 $\cos(2t) + \csc^{2}(t) = 0$
\end{problem}

\begin{problem}
$\tan^{3} \left( x \right) = 3\tan \left( x \right)$
\end{problem}

\begin{problem}
$\tan^{2} \left( \theta \right) = \frac{3}{2} \sec \left( \theta \right)$
\end{problem}

\begin{problem}
$\cos^{3} \left( t \right) = -\cos \left( t \right)$ 
\end{problem}

\begin{problem}
$\tan (2x) - 2\cos(x) = 0$
\end{problem}

\begin{problem}
$\csc^{3}(\theta) + \csc^{2}(\theta) = 4\csc(\theta) + 4$
\end{problem}

\begin{problem}
$2\tan(t) = 1 - \tan^{2}(t)$
\end{problem}

\begin{problem} \label{solveidentlast}
$\tan \left( x \right) = \sec \left( x \right)$ 
\end{problem}

\end{question}

\begin{question}
In Exercises \ref{solvemoreidentfirst} - \ref{solvemoreidentlast}, solve the equation, giving the exact solutions which lie in $[0, 2\pi)$

\begin{problem} \label{solvemoreidentfirst}
$\sin(6\theta) \cos(\theta) = -\cos(6\theta) \sin(\theta)$ 
\end{problem}

\begin{problem}
$\sin(3t)\cos(t) = \cos(3t) \sin(t)$
\end{problem}

\begin{problem}
$\cos(2x)\cos(x) + \sin(2x)\sin(x) = 1$ 
\end{problem}

 \begin{problem}
 $\cos(5\theta)\cos(3\theta) - \sin(5\theta)\sin(3\theta) = \frac{\sqrt{3}}{2}$
 \end{problem}

\begin{problem}
$\sin(t) + \cos(t) = 1$
\end{problem}

\begin{problem}
$\sin(x) + \sqrt{3} \cos(x) = 1$
\end{problem}

\begin{problem}
$\sqrt{2} \cos(\theta) - \sqrt{2} \sin(\theta) = 1$
\end{problem}

\begin{problem}
$\sqrt{3} \sin(2t) +  \cos(2t) = 1$
\end{problem}

\begin{problem}
 $\cos(2x) - \sqrt{3} \sin(2x) = \sqrt{2}$
\end{problem}
 
\begin{problem}
$3\sqrt{3}\sin(3\theta) - 3\cos(3\theta) = 3\sqrt{3}$
\end{problem}

\begin{problem}
$\cos(3t) = \cos(5t)$
\end{problem}

\begin{problem}
$\cos(4x) = \cos(2x)$
\end{problem}

\begin{problem}
$\sin(5\theta) = \sin(3\theta)$
\end{problem}

\begin{problem}
$\cos(5t) = -\cos(2t)$
\end{problem}

\begin{problem}
$\sin(6x) + \sin(x) = 0$
\end{problem}

\begin{problem} \label{solvemoreidentlast}
$\tan(x) = \cos(x)$ 
\end{problem}

\end{question}

\begin{question}
In Exercises \ref{firstinveqn} - \ref{lastinveqn}, solve the equation.

\begin{problem} \label{firstinveqn}
$\arccos(2x) = \pi$    %Ans $x = -\frac{1}{2}$
\end{problem}

\begin{problem}
$\pi - 2\arcsin(t) = 2\pi$   %Ans  $x=-1$
\end{problem}

\begin{problem}
$4\arctan(3x-1)-\pi=0$   %Ans $x = \frac{2}{3}$
\end{problem}

\begin{problem}
$6 \, \text{arccot}(2t) - 5\pi = 0$   %Ans  $x=-\frac{\sqrt{3}}{2}$
\end{problem}

\begin{problem}
$4 \,\text{arcsec}\left(\frac{x}{2}\right) = \pi$   %Ans $x = 2\sqrt{2}$
\end{problem}

\begin{problem}
$12 \,\text{arccsc}\left(\frac{t}{3}\right) = 2\pi$   %Ans $x = 6$
\end{problem}

\begin{problem}
$9 \arcsin^{2}(x) - \pi^2 = 0$   %Ans $x = \pm \frac{\sqrt{3}}{2}$
\end{problem}

\begin{problem}
$9 \arccos^{2}(t) - \pi^2 = 0$   %Ans $x = \frac{1}{2}$
\end{problem}

 \begin{problem}
 $8 \, \text{arccot}^{2}(x)+3\pi^2=10 \pi \, \text{arccot}(x)$   %Ans $x = -1,0$
\end{problem}
 
\begin{problem} \label{lastinveqn}
$6 \arctan(t)^2= \pi \arctan(x)+\pi^2$    %Ans $x = -\sqrt{3}$
\end{problem}

\end{question}

\begin{question}
In Exercises \ref{firstineqfirst} - \ref{firstineqlast}, solve the inequality.  Express the exact answer in \underline{interval} notation, restricting your attention to $0 \leq x \leq 2\pi$.

\begin{problem} \label{firstineqfirst}
$\sin \left( x \right) \leq 0$ 
\end{problem}

\begin{problem}
$\tan \left( t \right) \geq \sqrt{3}$
\end{problem}

\begin{problem}
$\sec^{2} \left( x \right) \leq 4$
\end{problem}

\begin{problem}
 $\cos^{2} \left( t \right) > \frac{1}{2}$
\end{problem}
 
\begin{problem}
$\cos \left( 2x \right) \leq 0$
\end{problem}

\begin{problem}
$\sin \left( t + \frac{\pi}{3} \right) > \frac{1}{2}$
\end{problem}

\begin{problem}
$\cot^{2} \left( x \right) \geq \frac{1}{3}$
\end{problem}

\begin{problem}
$2\cos(t) \geq 1$ \vphantom{$\frac{1}{2}$}
\end{problem}

\begin{problem}
$\sin(5x) \geq 5$ \vphantom{$\frac{1}{2}$}
\end{problem}

\begin{problem}
$\cos(3t) \leq 1$
\end{problem}

\begin{problem}
$\sec(x) \leq \sqrt{2}$
\end{problem}

\begin{problem} \label{firstineqlast}
$\cot(t) \leq 4$ 
\end{problem}

\end{question}

\begin{question}
In Exercises \ref{secondineqefirst} - \ref{secondineqlast}, solve the inequality.  Express the exact answer in \underline{interval} notation, restricting your attention to $-\pi \leq x \leq \pi$.

\begin{problem} \label{secondineqefirst}
$\cos \left( x \right) > \frac{\sqrt{3}}{2}$ 
\end{problem}

\begin{problem}
$\sin(t) > \frac{1}{3}$
\end{problem}

\begin{problem} 
$\sec \left( x \right) \leq 2$
\end{problem}

\begin{problem}
$\sin^{2} \left( t \right) < \frac{3}{4}$
\end{problem}

\begin{problem}
$\cot \left( x \right) \geq -1$
\end{problem}

\begin{problem} \label{secondineqlast}
$\cos(t) \geq \sin(t)$
\end{problem}

\end{question}

\begin{question}
In Exercises \ref{thirdineqfirst} - \ref{thirdineqlast}, solve the inequality.  Express the exact answer in \underline{interval} notation, restricting your attention to $-2\pi \leq x \leq 2\pi$.

\begin{problem} \label{thirdineqfirst}
$\csc \left( x \right) > 1$ 
\end{problem}

\begin{problem}
$\cos(t) \leq \frac{5}{3}$
\end{problem}

\begin{problem}
$\cot(x) \geq 5$
\end{problem}

\begin{problem}
$\tan^{2} \left( t \right) \geq 1$
\end{problem}

\begin{problem}
$\sin(2x) \geq \sin(x)$
\end{problem}

\begin{problem} \label{thirdineqlast}
$\cos(2t) \leq \sin(x)$
\end{problem}

\end{question}

\begin{question}
In Exercises \ref{invineqfirst} - \ref{invineqlast}, solve the given inequality.

\begin{problem}  \label{invineqfirst} 
$\arcsin(2x) > 0$
\end{problem}

\begin{problem}
$3 \arccos(t) \leq \pi$
\end{problem}

\begin{problem}
$6 \, \text{arccot}(7x) \geq \pi$ 
\end{problem}

\begin{problem}
$\pi > 2\arctan(t)$
\end{problem}

\begin{problem}
$2\arcsin(x)^2 > \pi \arcsin(x)$ 
\end{problem}

\begin{problem} \label{invineqlast} 
$12 \arccos(t)^2+2\pi^2>11\pi \arccos(t)$ 
\end{problem}

\end{question}

\begin{question}
In Exercises \ref{domainfirst} - \ref{domainlast}, express the domain of the function using the extended interval notation. (See Example \ref{TrigDomainEx1} and Section \ref{extendedinterval} for details.)

\begin{problem} \label{domainfirst}
$f(x) = \frac{1}{\cos(x) - 1}$
\end{problem}

\begin{problem}
$f(t) = \frac{\cos(t)}{\sin(t) + 1}$
\end{problem}

\begin{problem}
$f(x) = \sqrt{\tan^{2}(x) - 1}$
\end{problem}

\begin{problem}
$f(t) = \sqrt{2 - \sec(t)}$
\end{problem}

\begin{problem}
$f(x) = \csc(2x)$
\end{problem}

\begin{problem}
$f(t) = \frac{\sin(t)}{2 + \cos(t)}$
\end{problem}

\begin{problem}
$f(x) = 3\csc(x) + 4\sec(x)$ 
\end{problem}

\begin{problem}
$f(t) = \ln\left( |\cos(t)| \right)$
\end{problem}

\begin{problem}\label{domainlast}
$f(x) = \arcsin(\tan(x))$ 
\end{problem}

\end{question}

\begin{problem} \label{frequencynumberconnection} 

\begin{enumerate}

\item With the help of your classmates, determine the number of solutions to $\sin(x) = \frac{1}{2}$ in $[0,2\pi)$.  Then find the number of solutions to $\sin(2x) = \frac{1}{2}$,  $\sin(3x) = \frac{1}{2}$ and $\sin(4x) = \frac{1}{2}$ in $[0,2\pi)$. What pattern emerges?   Explain how this pattern would help you solve equations like $\sin(11x) = \frac{1}{2}$.  

\item Repeat the above exercise focusing on  $\sin\left(\frac{x}{2}\right)  = \frac{1}{2}$,  $\sin\left(\frac{3x}{2}\right)  = \frac{1}{2}$ and $\sin\left(\frac{5x}{2}\right)  = \frac{1}{2}$.  What pattern emerges here?  

\item  Replace sine with tangent and $\frac{1}{2}$ with $1$ and repeat the whole exploration.

\end{enumerate}

\end{problem}

\begin{problem}
Suppose an object weighing $10$ pounds is suspended from the ceiling by a spring which stretches $2$ feet to its equilibrium position when the object is attached.  

\begin{enumerate}

\item  Find the spring constant $k$ in $\frac{\text{lbs.}}{\text{ft.}}$ and the mass of the object in slugs.
\item  Find the equation of motion of the object if it is released from $1$ foot \textit{below} the equilibrium position from rest.  When is the first time the object passes through the equilibrium position? In which direction is it heading?
\item  Find the equation of motion of the object if it is released from $6$ inches \textit{above} the equilibrium position with a \textit{downward} velocity of $2$ feet per second.  Find when the object passes through the equilibrium position heading downwards for the third time.

\end{enumerate}

\end{problem}

\begin{problem} \label{sinecurvesketckex01IPs} 
In Example \ref{sinecurvesketckex01},  $f(x) = 2\sin(x) - \sin(2x)$  restricted to $0 \leq x \leq 2\pi$.  If $f''(x) = 4 \sin(2x) - 2\sin(x)$,  find the inflection points of the graph of $y = f(x)$.
\end{problem}

\begin{problem} \label{IPexamplecosineincdec}
In Example \ref{IPexamplecosine},  $f(x) =x-2\cos(x)$ restricted to $0 \leq x \leq 2\pi$.  If $f'(x) = 2\sin(x)+1$, list the open intervals over which $f$ is increasing and decreasing.  Find the local extrema. 
\end{problem}


\begin{problem} Let $f(x) = e^{-x} \, \sin(x)$ for $x > 0$.

\begin{enumerate}

\item  Use the Squeeze Theorem, Theorem \ref{squeezeth},  to find $\ds{\lim_{x \rightarrow \infty} f(x)}$.  Interpret your answer graphically.



\textbf{HINT:}   Since $-1 \leq \sin (x)  \leq 1$, $-e^{-x} \leq   e^{-x}   \sin (x)   \leq e^{-x}$ \ldots

\item  Use the fact that  $f'(x) = e^{-x} \cos(x) - e^{-x} \sin(x)$ to help you find the intervals over which $f$ is increasing and decreasing.

\item  Use the fact that $f''(x) = -2 e^{-x} \cos(x)$ to help you find the intervals over which the graph of $f$ is concave up and concave down.

\end{enumerate}

\end{problem}


\begin{problem} \label{squeezefordampedmotionexercise}
Recall  $x(t) = 10e^{-t/5} \sin\left(t + \frac{\pi}{3}\right)$ from  Example \ref{underdampedresonance} number \ref{underdampedproblem} models underdamped motion. Use the Squeeze Theorem, Theorem \ref{squeezeth}, to prove $\ds{\lim_{t \rightarrow \infty} x(t) = 0}$.

\textbf{HINT:}  Since $-1 \leq \sin\left(t + \frac{\pi}{3}\right) \leq 1$, $-10e^{-t/5} \leq   10e^{-t/5} \sin\left(t + \frac{\pi}{3}\right) \leq 10e^{-t/5}$ \ldots

\end{problem}

\end{document}


